The Design of an Aliasing-Tolerant Parallel Programming Model

### 2012-11-09 doc/types-pldi2013/alt_example.tex
updates

We begin by introducing an example program 
that motivates many of the features in Legion programming model
and provides an intuition for many of our design decisions.  The example program
executes a simulation of an arbitrary graph of electrical circuit elements.  
The simulation runs for many time steps and performs three computations 
for each time step: calculate new currents, distribute charge, and update voltages.
Due to space constraints we only introduce small code examples. 
The source code for the full circuit example written in the core Legion
language (see Section~\ref{sec:legioncore}) can be found in the
submitted appendix.

The simulation takes as input an arbitrary graph of circuit elements
that have been dynamically allocated into two logical regions: {\tt all\_nodes} 
and {\tt all\_wires}.  The types of nodes and wires are the tuples
shown in Listing~\ref{lst:tuples}.  The first two fields in the
{\tt CircuitWire} tuple are pointers that specify the two end points of
a wire.  Pointers are denoted by the @ symbol
followed by the region(s) into which the pointer may point.
The {\tt CircuitWire} type is parametrized by the local region
names {\tt rn} and {\tt rg} which contain {\tt CircuitNode} elements.
Our type system enforces the correctness of pointer types.

@@

The most important decision in any Legion program is the choice of
how to partition program data.  The ideal partitioning algorithm
depends on a multitude of factors including the type of data structure,
the input data, and the desired number of partitions (which often depends on the
size of the machine running the application).  Therefore
Legion permits applications to select their own partitioning algorithms.
In this example we invoke a third-party dynamic graph partitioning library
called METIS\cite{Metis98} as shown in Listing~\ref{lst:metis}.

@@

METIS takes as arguments lists of all the nodes and all the wires in
the graph.  It then breaks the graph into pieces by annotating each 
{\tt CircuitNode} with a piece ID (the last field in the tuple).
The function {\tt extern\_metis} is parameterized on the regions it will
access; in this case {\tt rn} is the region containing nodes, {\tt rw} is the region
contining wires, and {\tt rl} is the region containing the lists of pointers.  The function
must also specify the access {\em privileges} (e.g read, write) it requires for each of
these regions.  All three regions will be read by {\tt extern\_metis}, but only the {\tt rn}
region will be written to (since it will be modifying the last field of each {\tt CircuitNode}).
A function may only be invoked if the caller possesses all the privileges needed by the function
being called.  This is verified statically by the Legion type system.

@@

To efficiently support the access patterns of the circuit simulation,
multiple partitionings of the {\tt all\_nodes} region are performed.
The desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
First, we need subregions that describe the set of nodes ``owned'' by
each piece.  These will be called {\tt rn0}, {\tt rn1}, ... and since each
node has a single owner, they form the {\em disjoint} partition (indicated with a *)
on the left side of the tree.
Figure~\ref{sfig:part_fig:pieces} shows one possible output from
METIS with different colors identifing different piece IDs.
Second, each piece of the circuit will need to 
access the nodes that border its own.  These are commonly referred as {\em ghost nodes},
and the set of ghost nodes for the first two circuit pieces (named {\tt rg0} and {\tt rg1})
are shown in Figures~\ref{sfig:part_fig:ghost_zero}
and~\ref{sfig:part_fig:ghost_one}.  Because a node may neighbor more than one other circuit
piece, this second partition is an {\em aliased} partition.
Without the ability to have multiple partitionings of the same data, one would have two
possible alternatives:
\begin{enumerate}
\item Make a single partition with $2^N$ subregions, one for each possible case of sharing, or
\item Have all pieces use the {\tt all\_nodes} for their ghost cell region, greatly increasing the 
amount of data movement required
\end{enumerate}

To describe
the result of a partitioning algorithm, Legion uses {\em colorings}.
A coloring is a mapping from pointers to {\em colors} (represented in Core Legion as integers).  In
the circuit example, the piece IDs are the designated colors. 
Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
that constructs a coloring from a list of nodes which have been 
annotated with piece IDs.  The coloring of wires is similar, with each wire
using the color of its first endpoint.  For the ``ghost nodes'', a {\em multicoloring}
is used, which allows each pointer to be assigned multiple colorings.  (Multicolorings are
implemented in Core Legion with separate colorings for each color.)

\begin{lstlisting}[label={lst:coloring},caption={Coloring Construction}]
function create_piece_coloring[rl,rn](node_list: NodeList<rl,rn>@rl),
            reads(rl,rn) : coloring(rn) = 
    if isnull(node_list) then
        newcolor(rn) 
    else
        let list_elem : NodeList<rl,rn> = read(node_list) in
        let part_coloring : coloring(rn) = create_node_coloring(list_elem.2) in
        let node : CircuitNode = read(list_elem.1) in
            color(part_coloring, list_elem.1, node.5)
\end{lstlisting}

Colorings are the input for partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
partitioning.  The partitioning of wires is identical, while the partitioning
of the ghost nodes is performed in Core Legion as a separate partitioning operation
for each color.

@@

Since colorings can only color each pointer at most with a single color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions, each of which contains a subset of the original region.
The partition operation introduces constraints into the type system describing
both the disjointness of the subregions (e.g. {\tt rn0 $*$ rn1})
and the inclusion of each subregion in the original region
(e.g. {\tt rn0 $\leq$ all\_nodes}).

>>>>

We begin by introducing an example program 
motivating the novel features of Legion.
Due to space constraints we present only code fragments;
the full example is in the submitted appendix.

Consider a simulation of an electrical circuit.
The simulation runs for many time steps, each of which performs three computations:
 calculate new currents, distribute charges, and update voltages.
The simulation takes as input an arbitrary graph of circuit elements
that have been dynamically allocated in two logical regions: {\tt all\_nodes} 
and {\tt all\_wires}.  The types of nodes and wires are 
shown in Listing~\ref{lst:tuples}.  The first two fields in the
{\tt CircuitWire} tuple are pointers that specify the two end points of
a wire.  Pointers are denoted by the @ symbol
followed by the region(s) into which the pointer may point.
The {\tt CircuitWire} type is parametrized by the local region
names {\tt rn} and {\tt rg} which contain {\tt CircuitNode} elements.
Our type system enforces the correctness of pointer types.

@@

The most important decision in any Legion program is how to partition
program data.  An ideal partitioning depends on many factors,
including the shape of data structures, the input, and the number of
partitions, which usually varies with the target machine.  To allow
the necessary flexibility, in Legion partitioning is simply programmed,
rather than expressed in a more structured and restricted
sublanguage.  In the example we invoke METIS\cite{Metis98}, a dynamic graph
partitioning library called (see Listing~\ref{lst:metis}).

@@

METIS takes as arguments lists of the nodes and wires in
the graph.  It breaks the graph into pieces by annotating each 
{\tt CircuitNode} with a piece ID (the last field of the tuple).
The function {\tt extern\_metis} is parameterized on the regions it 
accesses: {\tt rn} is a region of nodes, {\tt rw} is a region of wires, 
and {\tt rl} is a region with the lists of pointers.  The function
also specifies the access {\em privileges} (e.g read, write) it requires for each 
region.  All three regions are read by {\tt extern\_metis}, but only the {\tt rn}
region is written (since it writes the piece ID).  A function can only 
have privileges that its caller also possesses; this property is verified statically. 
% by the Legion type system.

@@

To efficiently support the access patterns of the circuit simulation,
the {\tt all\_nodes} region is partitioned in two different ways.  The
desired {\em region tree} is shown in Figure~\ref{sfig:part_fig:tree}.
First, there are {\em subregions} of {\tt all\_nodes} that describe
the set of nodes ``owned'' by each piece, called {\tt rn0}, {\tt rn1},
$\ldots$. Since each node is in one piece, this partition is {\em
disjoint} (indicated by a $*$ on the left subtree).
Figure~\ref{sfig:part_fig:pieces} shows one possible node partition;
different colors identify different piece IDs.  Second, each piece of
the circuit needs access to the nodes on its border, commonly referred
to as {\em ghost nodes}.  The ghost nodes for two circuit pieces are
shown in Figures~\ref{sfig:part_fig:ghost_zero}
and~\ref{sfig:part_fig:ghost_one}; note two nodes are in both sets.
Because a node may neighbor more than one other circuit piece, this
second partition of {\tt all\_nodes} is {\em aliased}.  Thus, there
are two sources of aliasing in the region tree: the two distinct partitions
divide the {\tt all\_nodes} region in different ways, and the ghost
node subregions are not disjoint.

Note that there are two alternatives to using multiple partitions of the same data, both of which avoid introducing aliasing.
We could use a single partition with $2^n$ subregions, one for each
possible case of sharing; alternatively, all pieces could just use the
{\tt all\_nodes} region to access their ghost nodes.  Neither option
is attractive: the former significantly complicates programming, the
latter greatly increases data movement required at
runtime.

Programmers exoress partitions using {\em colorings},
maps from region elements to integers (``colors'').
Listing~\ref{lst:coloring} shows the function {\tt create\_piece\_coloring}
that constructs a coloring from the piece IDs of a list of nodes.
%The coloring of wires is similar, with each wire
%using the color of its first endpoint.  
For aliased partitions such as the ghost nodes, a {\em multicoloring}
is used, which assigns each region element (possibly) multiple colors.  
%(Multicolorings are
%implemented in Core Legion with separate colorings for each color.)
\begin{lstlisting}[label={lst:coloring},caption={Coloring Construction}]
function create_piece_coloring[rl,rn](node_list: NodeList<rl,rn>@rl),
            reads(rl,rn) : coloring(rn) = 
    if isnull(node_list) then
        newcolor(rn) 
    else
        let list_elem : NodeList<rl,rn> = read(node_list) in
        let part_coloring : coloring(rn) = create_node_coloring(list_elem.2) in
        let node : CircuitNode = read(list_elem.1) in
            color(part_coloring, list_elem.1, node.5)
\end{lstlisting}

Colorings are the input to partition operations in Legion.  The partition
statement shown in Listing~\ref{lst:partition} performs the ``owned nodes''
partitioning.    Since colorings assign each region element at most one color
(dynamically checked), the partition operation results in 
{\em disjoint} subregions, each a subset of the {\em parent} region.
A partition operation introduces constraints into the static type environment describing
both the disjointness of subregions (e.g., {\tt rn0 $*$ rn1})
and the subregion relationships
(e.g., {\tt rn0 $\leq$ all\_nodes}).  A partition created from a multicoloring
only introduces the subregion relationships, as the parition may be aliased.

%The partitioning of wires is identical, while the partitioning
%of the ghost nodes is performed in Core Legion as a separate partitioning operation
%for each color.

@@
